---
title: "Flexdashboard example - Leaflet Maps in a Storyboard"
output:   
    flexdashboard::flex_dashboard:
          storyboard: true
---

```{r setup, include=FALSE}

# Load necessary packages
# 1. Create vector - add any necessary package names needed.
list_of_packages <- c("leaflet", "leaflet.extras", "flexdashboard", "rgdal", "sf", "readr", 
                      "dplyr", "shiny", "htmltools", "plotly", "rmarkdown", "stringr", "viridis")
# 2. Create function that will do the job for us whenever required.
load_packages <- function(package_names){
  # Assign packages not listed in installed.packages() to a new vector object
  new_packages <- package_names[!(package_names %in% installed.packages()[,"Package"])]
  # If length of new_packages returns a value, install all packages named
  if(length(new_packages)) install.packages(new_packages, repos = "https://cloud.r-project.org/")
  # Use lappy to load required packages into the workspace
  lapply(package_names, library, character.only = TRUE)
}
# Use the new function to install and load necessary packages
load_packages(list_of_packages)


# Load Data in this chunk so it is available to all users.

# I mocked up dummy data for this example, which is supposed to represent very basic accounts data for a fictional
# business that has stores in different locations around Scotland.  There is one store per location and For each store 
# location, you get generic sales, costs, overheads, tax and profit data.  I've done this so you can see how each of the 
# different pieces of information could be shown on a map in a way that might be useful to a user.

# load in data as csv - I have made this data available to you so you can run this .rmd.
income <- read.csv(file = "income.csv", stringsAsFactors = FALSE, encoding = "latin1")

# load in boundaries - note that these are postcode boundaries I created in QGIS and are based on postocde areas.
# readOGR reads ESRI sahpefiles (.shp) and loads them as a SpatialDataFrame - if you are not familiar with these,
# see here: https://cengel.github.io/rspatial/2_spDataTypes.nb.html
shapes <- readOGR(dsn=getwd(), layer = "uk_pc_areas_simplified")

# Data cleansing (brief)

# No need for polygons for all of the UK as our fictional company only has stores in Scotland.  
# Create vector of required postcode areas for use in subsetting
pcode_sub <- c("AB", "DD", "DG", "EH", "FK", "G", "HS", "IV", "KA", "KW", "KY", "ML", "PA", "PH", "TD", "ZE")
# Subset shapes - with SpatialDataFrames, you should always subset and not filter as a filter create broken links between
# data in the slots, which R will automatically attempt to remedy and which will give you odd results.
shapes <- subset(shapes, shapes@data$PC_AREA %in% pcode_sub)
# Check only polygons for Scotland remain
# plot(shapes)
# PC_AREA required later but is a factor.  Change to char
shapes@data$PC_AREA <- as.character(shapes@data$PC_AREA)

# Ensure CRS is the same as our input data - 4326
shapes <- spTransform(shapes, CRS("+init=epsg:4326"))

```



```{r echo=FALSE}
# For this example I add each variable in the income data as a separate layer.  This is just to demonstrate
# how to add layers, how to use layer groups and to add a graduated colour scheme for each variable. 

## Polygon Layers and Layer  Goups

# leaflet makes great use of 'groups', which allow you to bundle together elements of your htmlwidget.  This comes
# in useful when you want to select specific layers on your map, linking elements such as layers and legends together
# and a load of other fun uses.

# Start by creating master polygons object that will be the basis for all our layers
master_poly <- shapes
# join income data from our .csv to the 'master_poly' data slot in the SpatialDataFrame and drop unnecessary columns
master_poly@data <- master_poly@data %>%
  left_join(income, by = c("PC_AREA" = "PostalArea")) %>%
  dplyr::select(-c("layer")) 


# Here we create color palettes that will be used to apply a gradient to each group.  I create a separate palette for 
# legends as the colours appear reversed when reordering the legend values.

pal1 <- colorBin("Blues", 
                     domain = master_poly@data$LocalPopulation, bins = 5, pretty = FALSE)
pal1A <- colorBin("Blues", reverse = TRUE,
                     domain = master_poly@data$LocalPopulation, bins = 5, pretty = FALSE)
# Our input data has '£,' in the values, which will not work well with colourBin (or similar functions).  So, we use
# stringr functions to remove them, then we use as.numeric to ensure the input to colorBin is numeric (not char).
pal2 <- colorBin("Blues", bins = 5, pretty = FALSE, 
                     domain = as.numeric(str_remove_all(master_poly@data$Sales, "[£,]" )))
pal2A <- colorBin("Blues", bins = 5, pretty = FALSE, reverse = TRUE,
                     domain = as.numeric(str_remove_all(master_poly@data$Sales, "[£,]" )))
# Notice houw you can add different colour scales to colour your gradients;
pal3 <- colorBin("Reds", bins = 5, pretty = FALSE, 
                     domain = as.numeric(str_remove_all(master_poly@data$Costs, "[£,]" )))
# But you must(!) remember to apply it to all elements in the same group for your map to make sense...
pal3A <- colorBin("Blues", bins = 5, pretty = FALSE, reverse = TRUE,
                     domain = as.numeric(str_remove_all(master_poly@data$Costs, "[£,]" )))
# Like blues, we use greens as a colour scale.  Be careful to pick colour scales that can cope with the number
# of bins you have - see RColorbrewer for more.  Also, make sure you have an appropriate number of bins or you 
# will get odd results.
pal4 <- colorBin("Greens", bins = 20, pretty = FALSE, 
                     domain = as.numeric(str_remove_all(master_poly@data$Overheads, "[£,]" )))
pal4A <- colorBin("Greens", bins = 20, pretty = FALSE, reverse = TRUE,
                     domain = as.numeric(str_remove_all(master_poly@data$Overheads, "[£,]" )))
# You can also use non-standard colour palettes, such as those available in the Viridis package.  Note that these
# are colourblind friendly, so may be of benefit if publishing your work...
pal5 <- colorBin("viridis", bins = 5, pretty = FALSE, 
                     domain = as.numeric(str_remove_all(master_poly@data$Tax, "[£,]" )))
pal5A <- colorBin("viridis", bins = 5, pretty = FALSE, reverse = TRUE,
                     domain = as.numeric(str_remove_all(master_poly@data$Tax, "[£,]" )))
# Using pretty=TRUE, R will determine the best number of bins to get 'pretty' breaks.  It will ignore
# the bins=5 and change it to the most appropriate number if it cannot give you pretty bins using your settings.
# Here I ask for six bins but 'pretty=TRUE' changes it to five;
pal6 <- colorBin("Blues", bins = 6, pretty = TRUE, 
                     domain = as.numeric(str_remove_all(master_poly@data$Profit, "[£,]" )))
pal6A <- colorBin("Blues", bins = 6, pretty = TRUE, reverse = TRUE,
                     domain = as.numeric(str_remove_all(master_poly@data$Profit, "[£,]" )))

# I create a different palette for each variable as the 'colourBin' function calculates how large
# each bin should be depending on the values in that specific column.  That way, each layer will have
# gradiated colours that give meaningful indications of value. 

```


Leaflet Maps {.storyboard}
===================================

### Local Population {data-commentary-width=400}

```{r map}

# Note from above that {.storyboard} is applied to this page on the dashboard.  This means that this
# page will be a storyboard.  Any other pages are not affected, so you can have a mix of layouts
# in a single flexdashboard.

# Also, we can specify the width of a commentary section that can be added to each element.  Below
# you will notice ***, which designates the start of a commentary section.  You can add basic text 
# immediately after this (ensuring one line of space) without having to write any further code.

## CREATE LEAFTLET MAP

# Create basic starting point, adding two potential basemaps.  Here I use OSM's standard map and 
# ESRI's grey canvas, which the user will be able to choose between in the finished product.

m <-  leaflet() %>%
  addTiles(group = "OSM") %>%
  addProviderTiles(provider = providers$Esri.WorldGrayCanvas, group = "ESRI Grey")


# Add each of the target vairables in our data as individual polygon layers on the basic map

# In these steps we use 'addPolygons' to add our shapes to the map, along with values from each
# target variable as separate layers.  

# In the addPolygons function we specify the input data, apply a group, give some styling to the
# polygon, including line colour, fill colour and transparency.  We then repeat this for when the
# polygon is highlighted, so that it appears differently than when not highlighted.  Next we 
# specify how to label each polygon in the layer - here I use values from the same variable but
# you are not restricted - and we specify content/formatting of a pop-up that appears when
# the polygon is clicked.
  
#Population
m <- m %>% addPolygons(data = master_poly, group = "LocalPopulation",
                  weight = 1, fillColor = ~pal1(LocalPopulation), opacity = 0.75,
                  stroke = TRUE, color = "Grey", dashArray = "3", fillOpacity = 0.7,
                  highlight = highlightOptions(
                    weight = 2,
                    fillOpacity = 0.9,
                    bringToFront = TRUE),
                  label = master_poly@data$LocalPopulation,
                  popup = ~paste0( "<h3>", Location, "</h3>", 
                                   "Population: ", prettyNum(LocalPopulation, big.mark = ","), "<br>",
                                   "Postcode Area: ", master_poly@data$PC_AREA)) %>%

  # We want every layer to have it's own legend and we want each legend to appear with the correct
  # set of polygons, so we add the legend to the same group.

        addLegend(position = "bottomright", pal = pal1A, values = rev(master_poly@data$LocalPopulation), 
                group = "LocalPopulation", title = "Local Population Count", opacity = 0.9,
                # here I create a function that will reorder the legend so the largest values are at the top
                labFormat = labelFormat(digits=-3, transform = function(x) sort(x, decreasing = TRUE))) %>%

# Now we repeat this process for all other variables that we want to use;
  
# Sales

      addPolygons(data = master_poly, group = "Sales",
            weight = 1, fillColor = ~pal2(as.numeric(str_remove_all(master_poly@data$Sales, "[£,]" ))), 
            opacity = 0.95, stroke = TRUE, color = "Grey", dashArray = "3", fillOpacity = 0.7,
            highlight = highlightOptions(
              weight = 2,
              fillOpacity = 0.9,
              bringToFront = TRUE),
            label = master_poly@data$Sales,
            popup = ~paste0( "<h3>", Location, "</h3>", 
                             "Sales: ", master_poly@data$Sales, "<br>", 
                             "Postcode Area: ", master_poly@data$PC_AREA)) %>%
      addLegend(position = "bottomright", pal = pal2A, values = master_poly@data$Sales,
                group = "Sales", title = "Sales", opacity = 0.9,
                labFormat = labelFormat(prefix = "£", digits=-2, transform = function(x) sort(x, decreasing = TRUE))) %>%

# Costs

      addPolygons(data = master_poly, group = "Costs",
            weight = 1, fillColor = ~pal3(as.numeric(str_remove_all(master_poly@data$Costs, "[£,]" ))), 
            opacity = 0.95, stroke = TRUE, color = "Grey", dashArray = "3", fillOpacity = 0.7,
            highlight = highlightOptions(
              weight = 2,
              fillOpacity = 0.9,
              bringToFront = TRUE),
            label = master_poly@data$Costs,
            popup = ~paste0( "<h3>", Location, "</h3>", 
                             "Costs: ", master_poly@data$Costs, "<br>", 
                             "Postcode Area: ", master_poly@data$PC_AREA)) %>%
      addLegend(position = "bottomright", pal = pal3A, values = master_poly@data$Costs,
                group = "Costs", title = "Costs", opacity = 0.9,
                labFormat = labelFormat(prefix = "£", digits=-2, transform = function(x) sort(x, decreasing = TRUE))) %>%

# Overheads

      addPolygons(data = master_poly, group = "Overheads",
            weight = 1, fillColor = ~pal4(as.numeric(str_remove_all(master_poly@data$Overheads, "[£,]" ))), 
            opacity = 0.95, stroke = TRUE, color = "Grey", dashArray = "3", fillOpacity = 0.7,
            highlight = highlightOptions(
              weight = 2,
              fillOpacity = 0.9,
              bringToFront = TRUE),
            label = master_poly@data$Overheads,
            popup = ~paste0( "<h3>", Location, "</h3>", 
                             "Overheads: ", master_poly@data$Overheads, "<br>", 
                             "Postcode Area: ", master_poly@data$PC_AREA)) %>%
      addLegend(position = "bottomright", pal = pal4A, values = master_poly@data$Overheads,
                group = "Overheads", title = "Overheads", opacity = 0.9,
                labFormat = labelFormat(prefix = "£", digits=-2, transform = function(x) sort(x, decreasing = TRUE))) %>%

# Tax

      addPolygons(data = master_poly, group = "Tax",
            weight = 1, fillColor = ~pal5(as.numeric(str_remove_all(master_poly@data$Tax, "[£,]" ))), 
            opacity = 0.95, stroke = TRUE, color = "Grey", dashArray = "3", fillOpacity = 0.7,
            highlight = highlightOptions(
              weight = 2,
              fillOpacity = 0.9,
              bringToFront = TRUE),
            label = master_poly@data$Tax,
            popup = ~paste0( "<h3>", Location, "</h3>", 
                             "Tax: ", master_poly@data$Tax, "<br>", 
                             "Postcode Area: ", master_poly@data$PC_AREA)) %>%
      addLegend(position = "bottomright", pal = pal5A, values = master_poly@data$Tax,
                # by removing 'digits=2' we notice that the legend contains decimals that we might not want
                group = "Tax", title = "Tax", opacity = 0.9,
                labFormat = labelFormat(prefix = "£", transform = function(x) sort(x, decreasing = TRUE))) %>%


# Profit 

      addPolygons(data = master_poly, group = "Profit",
            weight = 1, fillColor = ~pal6(as.numeric(str_remove_all(master_poly@data$Profit, "[£,]" ))), 
            opacity = 0.95, stroke = TRUE, color = "Grey", dashArray = "3", fillOpacity = 0.7,
            highlight = highlightOptions(
              weight = 2,
              fillOpacity = 0.9,
              bringToFront = TRUE),
            label = master_poly@data$Profit,
            popup = ~paste0( "<h3>", Location, "</h3>", 
                             "Profit: ", master_poly@data$Profit, "<br>", 
                             "Postcode Area: ", master_poly@data$PC_AREA)) %>%
      addLegend(position = "bottomright", pal = pal6A, values = master_poly@data$Profit,
                group = "Profit", title = "Profit", opacity = 0.9,
                labFormat = labelFormat(prefix = "£", digits=-2, transform = function(x) sort(x, decreasing = TRUE))) %>%


# Add layer controls
# Finally we need to add some controls so that our map behaves in the way we want.  
  # Basegroups give the user radial boxes in which only one group of elements can be selected (e.g. basemaps)
  # Overlaygroups give the user checkboxes that allow selection of multiple elements (e.g. layers)
  # In the options I specify that the container for the groups are not collapsed when the document opens
  # Hidegroup lets you specify which elements should not be automatically selected when the document opens.

     addLayersControl( baseGroups = c("OSM", "ESRI Grey"),  
                            overlayGroups = c("LocalPopulation", "Sales", "Costs", "Overheads", "Tax", "Profit"),
                            options = layersControlOptions(collapsed = FALSE)) %>%
    hideGroup(c("Sales", "Costs", "Overheads", "Tax", "Profit"))


m
```


***
Commentary

This is our commentary section where you can add simple, unformatted text to your map.  As standard, this section is set to a width of 250 but I tailored this to 400. In flexdashboard, you have a range of options for setting sizing (see [here](https://rmarkdown.rstudio.com/flexdashboard/using.html#sizing)).

Empty lines are accepted in your Commentary section but other items may be ignored if you are not careful, such as a random linebreak in a sentence;

If you want to add more detail you can use basic
markdown tools like **bold** or *italic*.  Or you can write HTML that will further customise your text using the HTML function;

```{r Comments}
HTML("<big><strong> This is a Title </strong></big>
      <br>
      <p>By mixing in HTML tags, you can tailor the commentary section, which otherwise would be simple text.</p>
      <center><p> You can also include HTML CSS that will make the buttons smaller in storyboard layout </p></center>.
      ")
```

You can even add bulleted lists...

* Item one
    + sub one

And ordered lists... 

1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b
    
But be careful as some markdown formatting will conflict with flexdashbaord.  For example, this commentary section will continue until a 
new page (=======) or container/section (###) is specified.  The specifier (###) in flexdashboard is also a heading specifier 
in markdown, so R gets confused if you add #s at the start of a line in storyboard commentary.

```{r}
# Finally, the below code appears at the bottom of this example and it applies CSS styling to the buttons at the top of the storyboard.
# Try adding/removing it to notice the differences it makes.
# <style>
# 
# .storyboard-nav .sbframelist {
#     margin: 0 auto;
#     width: 94%;
#     height: 50px;
#     overflow: hidden;
#     text-shadow: none;
#     margin-bottom: 8px;
# }
# 
# .storyboard-nav .sbnext, .storyboard-nav .sbprev {
#     width: 2%;
#     height: 50px;
#     font-size: 25px;
# }
# 
# </style>
```



**I hope you find this useful!**  

Anthony






<style>

.storyboard-nav .sbframelist {
    margin: 0 auto;
    width: 94%;
    height: 50px;
    overflow: hidden;
    text-shadow: none;
    margin-bottom: 8px;
}

.storyboard-nav .sbnext, .storyboard-nav .sbprev {
    width: 2%;
    height: 50px;
    font-size: 25px;
}

</style>
